---
title: "Outstanding EDA: Global CO2 Emissions Panel"
author: "Statistical Computing Project"
date: "`r format(Sys.Date(), '%Y-%m-%d')`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    df_print: paged
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.width = 10,
  fig.height = 6,
  dpi = 150
)
```

## 1) Libraries and helpers

```{r libraries}
library(tidyverse)
library(scales)
library(corrplot)

if (!dir.exists("output")) dir.create("output", recursive = TRUE)
if (!dir.exists("output/figures")) dir.create("output/figures", recursive = TRUE)
if (!dir.exists("output/tables")) dir.create("output/tables", recursive = TRUE)

theme_set(
  theme_minimal(base_size = 13) +
    theme(
      plot.title = element_text(face = "bold", size = 15),
      plot.subtitle = element_text(size = 11),
      axis.title = element_text(face = "bold"),
      panel.grid.minor = element_blank()
    )
)

save_plot <- function(plot_obj, filename, width = 10, height = 6) {
  ggsave(
    filename = file.path("output/figures", filename),
    plot = plot_obj,
    width = width,
    height = height,
    dpi = 150
  )
}
```

## 2) Load data and validate schema

```{r load-data}
co2 <- read_csv("cleaned_co2_data_20vars.csv", show_col_types = FALSE)

expected_cols <- c(
  "country", "year", "iso_code", "population", "gdp",
  "primary_energy_consumption", "energy_per_capita", "energy_per_gdp",
  "co2", "co2_per_capita", "total_ghg", "consumption_co2",
  "methane", "nitrous_oxide", "cement_co2", "flaring_co2", "trade_co2",
  "coal_co2", "gas_co2", "oil_co2"
)

stopifnot(all(expected_cols %in% names(co2)))
stopifnot(ncol(co2) == length(expected_cols))

glimpse(co2)
```

```{r dataset-at-a-glance}
data_snapshot <- tibble(
  n_rows = nrow(co2),
  n_columns = ncol(co2),
  n_countries = n_distinct(co2$country),
  first_year = min(co2$year, na.rm = TRUE),
  last_year = max(co2$year, na.rm = TRUE)
)

knitr::kable(data_snapshot, caption = "Dataset at a glance")
write_csv(data_snapshot, "output/tables/eda_data_snapshot.csv")
```

## 3) Data quality checks

```{r quality-checks}
missing_summary <- co2 %>%
  summarise(across(everything(), ~ sum(is.na(.x)))) %>%
  pivot_longer(everything(), names_to = "variable", values_to = "missing_count") %>%
  mutate(missing_pct = missing_count / nrow(co2))

knitr::kable(
  missing_summary %>% arrange(desc(missing_count)),
  caption = "Missing values by variable"
)

duplicate_rows <- sum(duplicated(co2))
cat("Duplicate rows:", duplicate_rows, "\n")
```

## 4) Univariate profile of key variables

```{r numeric-summary}
numeric_summary <- co2 %>%
  summarise(
    across(
      where(is.numeric),
      list(
        min = ~ min(.x, na.rm = TRUE),
        p25 = ~ quantile(.x, 0.25, na.rm = TRUE),
        median = ~ median(.x, na.rm = TRUE),
        mean = ~ mean(.x, na.rm = TRUE),
        p75 = ~ quantile(.x, 0.75, na.rm = TRUE),
        max = ~ max(.x, na.rm = TRUE),
        sd = ~ sd(.x, na.rm = TRUE)
      ),
      .names = "{.col}__{.fn}"
    )
  ) %>%
  pivot_longer(
    cols = everything(),
    names_to = c("variable", "stat"),
    names_sep = "__",
    values_to = "value"
  ) %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  arrange(variable)

knitr::kable(numeric_summary, digits = 3, caption = "Numerical summary statistics")
write_csv(numeric_summary, "output/tables/eda_numeric_summary.csv")
```

```{r target-distribution}
p_target <- ggplot(co2, aes(x = co2)) +
  geom_histogram(bins = 45, fill = "#0B7285", color = "white", alpha = 0.9) +
  scale_x_continuous(labels = label_number_si()) +
  labs(
    title = "Distribution of Total CO2 Emissions",
    subtitle = "Strong right skew is expected in country-level emissions",
    x = "CO2 emissions",
    y = "Count"
  )

save_plot(p_target, "eda_target_distribution.png")
p_target
```

```{r target-distribution-log}
p_target_log <- ggplot(co2, aes(x = log1p(co2))) +
  geom_histogram(bins = 45, fill = "#F08C00", color = "white", alpha = 0.9) +
  labs(
    title = "Log-Transformed CO2 Distribution",
    subtitle = "Log transform reveals the body of the distribution clearly",
    x = "log(1 + CO2)",
    y = "Count"
  )

save_plot(p_target_log, "eda_target_distribution_log.png")
p_target_log
```

## 5) Time dynamics and concentration patterns

```{r global-trend}
global_trend <- co2 %>%
  group_by(year) %>%
  summarise(
    global_co2 = sum(co2, na.rm = TRUE),
    median_country_co2 = median(co2, na.rm = TRUE),
    q10_country_co2 = quantile(co2, 0.10, na.rm = TRUE),
    q90_country_co2 = quantile(co2, 0.90, na.rm = TRUE),
    .groups = "drop"
  )

p_global <- ggplot(global_trend, aes(x = year)) +
  geom_ribbon(
    aes(ymin = q10_country_co2, ymax = q90_country_co2),
    fill = "#A5D8FF", alpha = 0.35
  ) +
  geom_line(aes(y = median_country_co2, color = "Median country CO2"), linewidth = 1.1) +
  geom_line(aes(y = global_co2 / 100, color = "Global total CO2 / 100"), linewidth = 1.1) +
  scale_color_manual(values = c("Median country CO2" = "#0C8599", "Global total CO2 / 100" = "#C92A2A")) +
  scale_y_continuous(labels = label_number_si()) +
  labs(
    title = "CO2 Through Time: Global Growth and Country Dispersion",
    subtitle = "Ribbon shows country 10th to 90th percentile range each year",
    x = "Year",
    y = "Emissions (scaled)",
    color = NULL
  )

save_plot(p_global, "eda_global_trend.png")
p_global
```

```{r top-emitters-latest-year}
latest_year <- max(co2$year, na.rm = TRUE)

top_emitters <- co2 %>%
  filter(year == latest_year) %>%
  arrange(desc(co2)) %>%
  slice_head(n = 15) %>%
  mutate(country = fct_reorder(country, co2))

p_top <- ggplot(top_emitters, aes(x = country, y = co2)) +
  geom_col(fill = "#2B8A3E", alpha = 0.9) +
  coord_flip() +
  scale_y_continuous(labels = label_number_si()) +
  labs(
    title = paste("Top 15 Emitters in", latest_year),
    subtitle = "Concentration among a small group of countries is substantial",
    x = NULL,
    y = "CO2 emissions"
  )

save_plot(p_top, "eda_top_emitters_latest_year.png")
p_top
```

## 6) Relationship diagnostics

```{r gdp-and-energy-relationships}
co2_rel <- co2 %>%
  mutate(
    gdp_per_capita = gdp / population,
    co2_intensity = co2 / pmax(gdp, 1e-9)
  )

p_gdp_pc <- ggplot(co2_rel, aes(x = gdp_per_capita, y = co2_per_capita)) +
  geom_point(alpha = 0.35, color = "#1971C2") +
  geom_smooth(method = "loess", se = FALSE, color = "#0B7285", linewidth = 1.1) +
  scale_x_continuous(trans = "log10", labels = label_number_si()) +
  scale_y_continuous(labels = label_number(accuracy = 0.1)) +
  labs(
    title = "Economic Development and CO2 per Capita",
    subtitle = "Nonlinear pattern: emissions rise with income, then flatten for high-income countries",
    x = "GDP per capita (log scale)",
    y = "CO2 per capita"
  )

save_plot(p_gdp_pc, "eda_gdp_per_capita_vs_co2_per_capita.png")
p_gdp_pc
```

```{r energy-efficiency}
p_efficiency <- ggplot(co2_rel, aes(x = energy_per_gdp, y = co2_per_capita)) +
  geom_point(alpha = 0.35, color = "#5F3DC4") +
  geom_smooth(method = "lm", se = TRUE, color = "#3B5BDB", linewidth = 1.0) +
  scale_x_continuous(labels = label_number_si()) +
  labs(
    title = "Energy Intensity vs CO2 per Capita",
    subtitle = "Higher energy needed per GDP unit is generally associated with higher CO2 per capita",
    x = "Energy per GDP",
    y = "CO2 per capita"
  )

save_plot(p_efficiency, "eda_energy_per_gdp_vs_co2_per_capita.png")
p_efficiency
```

## 7) Source composition and structural shifts

```{r source-shares-over-time}
source_mix <- co2 %>%
  group_by(year) %>%
  summarise(
    coal = sum(coal_co2, na.rm = TRUE),
    gas = sum(gas_co2, na.rm = TRUE),
    oil = sum(oil_co2, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(cols = c(coal, gas, oil), names_to = "source", values_to = "co2_from_source") %>%
  group_by(year) %>%
  mutate(source_share = co2_from_source / sum(co2_from_source)) %>%
  ungroup()

p_mix <- ggplot(source_mix, aes(x = year, y = source_share, fill = source)) +
  geom_area(alpha = 0.92, color = "white", linewidth = 0.15) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c(coal = "#343A40", gas = "#339AF0", oil = "#F08C00")) +
  labs(
    title = "Global Fossil CO2 Composition Over Time",
    subtitle = "Shares from coal, gas, and oil (aggregated across countries)",
    x = "Year",
    y = "Share of fossil-source CO2",
    fill = "Source"
  )

save_plot(p_mix, "eda_source_composition_area.png")
write_csv(source_mix, "output/tables/eda_source_mix_by_year.csv")
p_mix
```

## 8) Correlation structure and multicollinearity signals

```{r correlation}
numeric_data <- co2 %>% select(where(is.numeric))
corr_mat <- cor(numeric_data, use = "pairwise.complete.obs")

corrplot(
  corr_mat,
  method = "color",
  type = "lower",
  order = "hclust",
  tl.col = "black",
  tl.cex = 0.7,
  addCoef.col = "black",
  number.cex = 0.5
)

write_csv(
  as_tibble(as.table(corr_mat), .name_repair = "minimal") %>%
    rename(var1 = Var1, var2 = Var2, corr = n),
  "output/tables/eda_correlation_long.csv"
)
```

```{r strongest-correlations}
strong_corr <- as_tibble(as.table(corr_mat), .name_repair = "minimal") %>%
  rename(var1 = Var1, var2 = Var2, corr = n) %>%
  filter(var1 != var2) %>%
  mutate(pair = if_else(var1 < var2, paste(var1, var2, sep = " | "), paste(var2, var1, sep = " | "))) %>%
  group_by(pair) %>%
  summarise(var1 = first(var1), var2 = first(var2), corr = first(corr), .groups = "drop") %>%
  mutate(abs_corr = abs(corr)) %>%
  arrange(desc(abs_corr)) %>%
  slice_head(n = 15)

knitr::kable(strong_corr, digits = 3, caption = "Top 15 absolute correlations")
write_csv(strong_corr, "output/tables/eda_top_correlations.csv")
```

## 9) Country-level risk and intensity diagnostics

```{r country-intensity}
country_profile <- co2_rel %>%
  group_by(country, iso_code) %>%
  summarise(
    avg_co2 = mean(co2, na.rm = TRUE),
    avg_co2_per_capita = mean(co2_per_capita, na.rm = TRUE),
    avg_co2_intensity = mean(co2_intensity, na.rm = TRUE),
    years_observed = n(),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_co2_per_capita))

knitr::kable(head(country_profile, 15), digits = 3, caption = "Top 15 countries by average CO2 per capita")
write_csv(country_profile, "output/tables/eda_country_profile.csv")
```

## 10) EDA conclusions for modeling decisions

1. **Heavy skew and scale heterogeneity:** `co2`, `gdp`, and source-specific emissions are highly right-skewed. Standardization (already used in the notebook) is essential; log features may improve linear models further.
2. **Strong multicollinearity:** Emission family variables are strongly correlated, which can destabilize OLS coefficients. This justifies Ridge/Lasso/Elastic Net and dimensionality reduction with PCA.
3. **Concentration and temporal dynamics:** A small set of countries dominates total emissions, while distribution across countries remains broad over time. Model validation should remain time-aware.
4. **Energy-economic links are nonlinear:** Relationships such as GDP-per-capita vs CO2-per-capita are curved, supporting nonlinear models (NN/LSTM) in addition to linear baselines.
5. **Composition shifts matter:** Coal, gas, and oil shares evolve over time, indicating structural changes that sequence-aware models can exploit.

---

### Reproducibility note

All EDA outputs are saved into:

- `output/figures/eda_*.png`
- `output/tables/eda_*.csv`

Knit this report from the repository root so paths resolve correctly.
