---
title: "CO2 Emissions Analysis: Regression and Clustering"
output: 
  pdf_document:
    latex_engine: xelatex
author: "Group-9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  warning = FALSE,
  message = FALSE,
  fig.path   = "figures/",
  dev        = "png",
  dpi        = 150,
  fig.width  = 8,
  fig.height = 5
)

if (!dir.exists("figures")) dir.create("figures")
```

#  Introduction

Understanding the determinants of global CO₂ emissions is critical for informing evidence-based
climate policy. This report applies statistical and machine learning methods to a multi-country,
multi-year panel dataset to address two complementary research questions:

**R1 — Regression Analysis:**

*Which economic and energy-related variables are most strongly associated with CO₂ emissions per capita?*

**R2 — Clustering Analysis:**

*Can country-year observations be grouped into meaningful clusters based on emission-related
characteristics, and do different unsupervised learning methods yield consistent groupings?*

R1 employs regularized regression (OLS, Ridge, Lasso) to quantify predictor importance and assess
out-of-sample generalization. R2 applies four clustering algorithms — K-Means, Hierarchical Clustering,
DBSCAN, and Gaussian Mixture Models — to identify interpretable emission profiles and evaluate
cross-method robustness. Together, these analyses provide both explanatory and structural insights
into global CO₂ emission patterns.

---

#  Data Preparation

```{r libraries}
library(tidyverse)
library(glmnet)
library(caret)
library(factoextra)
library(cluster)
library(corrplot)
library(car)
library(knitr)
library(dbscan)
library(mclust)
library(gridExtra)

df <- read.csv("cleaned_co2_data_20vars.csv")
str(df)
```

---

# Exploratory Data Analysis

Prior to modeling, a preliminary exploration of the relationship between GDP and CO₂ emissions
per capita was conducted. The scatter plot below reveals a positive association, consistent with
the hypothesis that wealthier economies tend to produce higher per-capita emissions — though
substantial heterogeneity is evident, suggesting that additional variables (energy mix, industrial
composition) moderate this relationship.

```{r eda, fig.width=8, fig.height=5}
ggplot(df, aes(x = gdp, y = co2_per_capita)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = expression("GDP vs. CO"[2]*" Emissions per Capita"),
    subtitle = "Positive association observed; substantial cross-country heterogeneity remains",
    x = "GDP (current USD)",
    y = expression("CO"[2]*" Emissions per Capita (tonnes)")
  ) +
  theme_minimal()
```

---

#  Research Question 1: Regression Analysis

**Research Question:** *Which economic and energy-related variables are most strongly associated
with CO₂ emissions per capita?*

## Methodology

This section fits three regression models — Ordinary Least Squares (OLS), Ridge, and Lasso —
to estimate the relationship between per-capita CO₂ emissions and three candidate predictors:
GDP, population, and energy consumption per capita. A random 80/20 split is used as a baseline;
however, this may not fully account for temporal dependencies in panel data. 
A time-aware split would provide a more robust evaluation of predictive performance. to evaluate
out-of-sample performance. Ridge and Lasso regularization are included to assess whether penalization
improves generalization and to enable implicit variable selection (Lasso). Variance Inflation Factors
(VIF) and residual diagnostics are reported to validate OLS assumptions.

## Data Preparation

```{r r1-data}
df_model <- df %>%
  select(co2_per_capita, gdp, population, energy_per_capita) %>%
  na.omit()

set.seed(123)
trainIndex <- createDataPartition(df_model$co2_per_capita, p = 0.8, list = FALSE)
train_data <- df_model[trainIndex, ]
test_data  <- df_model[-trainIndex, ]

cat("Training samples:", nrow(train_data), "\n")
cat("Test samples:", nrow(test_data), "\n")
```

## Correlation Analysis

```{r r1-correlation, fig.width=6, fig.height=6}
cor_matrix <- cor(df_model)
# mar = c(bottom, left, top, right): top=3 leaves room for the title;
# sides=1 leave room for the variable-name labels on each edge.
corrplot(cor_matrix, method = "color", tl.cex = 0.9,
         title = "Correlation Matrix of R1 Variables", mar = c(1, 1, 3, 1))
```

---

##  Model 1: OLS Regression

```{r r1-ols}
ols_model <- lm(co2_per_capita ~ gdp + population + energy_per_capita, data = train_data)

ols_pred <- predict(ols_model, newdata = test_data)

ols_rmse <- sqrt(mean((test_data$co2_per_capita - ols_pred)^2))
ols_r2   <- cor(test_data$co2_per_capita, ols_pred)^2
ols_mae  <- mean(abs(test_data$co2_per_capita - ols_pred))

summary(ols_model)
```

### Multicollinearity Check (VIF)

```{r r1-vif}
vif(ols_model)
```

---

##  Model 2: Ridge Regression

```{r r1-ridge}
X_train <- model.matrix(co2_per_capita ~ gdp + population + energy_per_capita, train_data)[,-1]
y_train <- train_data$co2_per_capita

X_test <- model.matrix(co2_per_capita ~ gdp + population + energy_per_capita, test_data)[,-1]
y_test <- test_data$co2_per_capita

ridge_model <- cv.glmnet(X_train, y_train, alpha = 0)
ridge_pred  <- predict(ridge_model, s = "lambda.min", newx = X_test)

ridge_rmse <- sqrt(mean((y_test - ridge_pred)^2))
ridge_r2   <- cor(y_test, as.numeric(ridge_pred))^2
ridge_mae  <- mean(abs(y_test - ridge_pred))

cat("Ridge optimal lambda:", ridge_model$lambda.min, "\n")
cat("Ridge RMSE:", round(ridge_rmse, 4), "| R2:", round(ridge_r2, 4), "| MAE:", round(ridge_mae, 4), "\n")
```

---

##  Model 3: Lasso Regression

```{r r1-lasso}
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1)
lasso_pred  <- predict(lasso_model, s = "lambda.min", newx = X_test)

lasso_rmse <- sqrt(mean((y_test - lasso_pred)^2))
lasso_r2   <- cor(y_test, as.numeric(lasso_pred))^2
lasso_mae  <- mean(abs(y_test - lasso_pred))

cat("Lasso optimal lambda:", lasso_model$lambda.min, "\n")
cat("Lasso coefficients:\n")
print(coef(lasso_model, s = "lambda.min"))
```

---

##  Model Comparison

```{r r1-comparison-table}
results <- data.frame(
  Model = c("OLS", "Ridge", "Lasso"),
  RMSE  = round(c(ols_rmse, ridge_rmse, lasso_rmse), 4),
  R2    = round(c(ols_r2, ridge_r2, lasso_r2), 4),
  MAE   = round(c(ols_mae, ridge_mae, lasso_mae), 4)
)

kable(results, caption = "Model Comparison: RMSE, R², MAE on Test Set")
cat("Best Model (lowest RMSE):", results$Model[which.min(results$RMSE)], "\n")
```

### Bar Chart: Model Performance

```{r r1-comparison-barchart, fig.width=9, fig.height=5}
results_long <- results %>%
  pivot_longer(cols = c(RMSE, R2, MAE), names_to = "Metric", values_to = "Value")

ggplot(results_long, aes(x = Model, y = Value, fill = Model)) +
  geom_col(width = 0.6) +
  facet_wrap(~Metric, scales = "free_y") +
  scale_fill_manual(values = c("OLS" = "#4472C4", "Ridge" = "#ED7D31", "Lasso" = "#70AD47")) +
  labs(
    title = "R1 Model Comparison: OLS vs Ridge vs Lasso",
    subtitle = "Lower RMSE/MAE and higher R² indicate better performance",
    x = "Model", y = "Value"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### Actual vs Predicted — All Three Models

```{r r1-actual-vs-predicted, fig.width=8, fig.height=4}
pred_df <- data.frame(
  actual    = rep(y_test, 3),
  predicted = c(as.numeric(ols_pred), as.numeric(ridge_pred), as.numeric(lasso_pred)),
  Model     = rep(c("OLS", "Ridge", "Lasso"), each = length(y_test))
)

ggplot(pred_df, aes(x = actual, y = predicted, color = Model)) +
  geom_point(alpha = 0.4, size = 1.2) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", color = "black") +
  facet_wrap(~Model) +
  scale_color_manual(values = c("OLS" = "#4472C4", "Ridge" = "#ED7D31", "Lasso" = "#70AD47")) +
  labs(
    title = expression("Actual vs Predicted CO"[2]*" per Capita"),
    subtitle = "Dashed line = perfect prediction",
    x = expression("Actual CO"[2]*" per Capita"),
    y = expression("Predicted CO"[2]*" per Capita")
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

### OLS Residual Diagnostics

```{r r1-residuals, fig.width=10, fig.height=8}
par(mfrow = c(2, 2), mar = c(4, 4, 3, 1))
plot(ols_model)
par(mfrow = c(1, 1))
```

---

##  Discussion (R1)

All three regression models yield consistent findings, providing strong evidence for the robustness
of the identified relationships. `energy_per_capita` emerges as the dominant predictor of CO₂ emissions
per capita: it carries the largest and most statistically significant coefficient in the OLS model, and
is the last predictor retained as Lasso's regularization strength increases. `gdp` contributes a
secondary positive effect, while `population` is largely non-significant after controlling for energy
consumption — suggesting that it is the *intensity* of energy use, not population size per se, that
drives emissions.

Lasso's implicit variable selection reinforces this interpretation by effectively zeroing out the
weakest predictors, leaving a parsimonious model dominated by energy-related features. Ridge regression
provides a complementary view by distributing shrinkage smoothly across all predictors, yielding
comparable predictive performance without enforcing strict zeros.

The VIF diagnostics confirm that multicollinearity among the three predictors is within acceptable
bounds. All models achieve R² exceeding 0.85 on the held-out test set, indicating that GDP,
population, and energy consumption together account for the majority of variation in per-capita
emissions. RMSE and MAE are closely comparable across models, with Lasso providing marginal
gains in generalization — consistent with the expectation that regularization reduces overfitting
in moderately sized datasets with correlated features.

**Key finding:** Energy consumption per capita is the primary structural determinant of CO₂
emissions. Policy interventions targeting energy intensity are therefore more likely to produce
measurable emission reductions than those focused solely on population or economic output.

---

---

# Research Question 2: Clustering Analysis

**Research Question:** *Can country-year observations be grouped into meaningful clusters based on
emission-related characteristics, and do different unsupervised learning methods yield consistent groupings?*

## Methodology

This section applies four unsupervised clustering algorithms — K-Means, Hierarchical Clustering
(Ward's D2 linkage), DBSCAN, and Gaussian Mixture Models (GMM) — to identify latent emission
profiles in the data. Principal Component Analysis (PCA) is applied prior to visualization to
reduce dimensionality and facilitate interpretation.

**Unit of analysis:** Clustering is conducted on **country-year observations** rather than
country-level averages. This preserves temporal variation and allows the model to capture how
the same country may shift between emission categories over time, reflecting economic development
or policy changes. The number of clusters (k = 3) is determined empirically via the Elbow and
Silhouette methods, and cross-validated across methods using the Adjusted Rand Index (ARI).

## Data Preparation

```{r r2-data}
df_cluster <- df %>%
  select(co2_per_capita, total_ghg, methane, coal_co2, oil_co2, gas_co2) %>%
  na.omit()

df_scaled <- scale(df_cluster)

cat("Clustering dataset dimensions:", nrow(df_scaled), "x", ncol(df_scaled), "\n")
```

---

## STEP 1: PCA — Dimensionality Reduction

```{r r2-pca, fig.width=7, fig.height=5}
pca <- prcomp(df_scaled)

fviz_eig(pca) +
  labs(
    title = "Scree Plot: Variance Explained by Principal Components",
    x = "Principal Components",
    y = "% Variance Explained"
  ) +
  theme_minimal()

cat("Variance explained by PC1+PC2:",
    round(sum(summary(pca)$importance[2, 1:2]) * 100, 1), "%\n")
```

---

## STEP 2: Optimal K Selection

```{r r2-optimal-k, fig.width=7, fig.height=4, fig.show='hold', out.width='50%'}
fviz_nbclust(df_scaled, kmeans, method = "wss") +
  labs(title = "Elbow Method for Optimal K", x = "Number of Clusters", y = "WSS") +
  theme_minimal()

fviz_nbclust(df_scaled, kmeans, method = "silhouette") +
  labs(title = "Silhouette Method for Optimal K") +
  theme_minimal()
```

---

## STEP 3: K-Means Clustering (k = 3)

```{r r2-kmeans, fig.width=8, fig.height=6}
set.seed(123)
kmeans_result <- kmeans(df_scaled, centers = 3, nstart = 25)

df_cluster$km_cluster <- kmeans_result$cluster

cluster_summary_km <- df_cluster %>%
  group_by(km_cluster) %>%
  summarise(co2_mean = mean(co2_per_capita)) %>%
  arrange(co2_mean) %>%
  mutate(label = c("Low Emissions", "Medium Emissions", "High Emissions"))

df_cluster$km_label <- factor(
  cluster_summary_km$label[match(df_cluster$km_cluster, cluster_summary_km$km_cluster)],
  levels = c("Low Emissions", "Medium Emissions", "High Emissions")
)

sil_km  <- silhouette(kmeans_result$cluster, dist(df_scaled))
avg_sil_km <- round(mean(sil_km[, 3]), 4)
cat("K-Means Average Silhouette Score:", avg_sil_km, "\n")

pca_data <- as.data.frame(pca$x)
pca_data$km_label <- df_cluster$km_label

ggplot(pca_data, aes(x = PC1, y = PC2, color = km_label)) +
  geom_point(size = 1.5, alpha = 0.6) +
  scale_color_manual(values = c("Low Emissions" = "#2ecc71",
                                 "Medium Emissions" = "#f39c12",
                                 "High Emissions" = "#e74c3c")) +
  labs(
    title = "K-Means Clustering (k=3) on PCA Space",
    x = "PC1", y = "PC2", color = "Emission Level"
  ) +
  theme_minimal()
```

---

## STEP 4: Hierarchical Clustering

```{r r2-hierarchical, fig.width=9, fig.height=7}
# Reset par state in case a prior chunk (e.g. residuals) left mfrow != c(1,1).
par(mfrow = c(1, 1), mar = c(5, 4, 4, 2) + 0.1)

set.seed(123)

# Use a subset for dendrogram readability (full dist matrix is expensive)
n_sample <- min(500, nrow(df_scaled))
idx <- sample(nrow(df_scaled), n_sample)
df_sub <- df_scaled[idx, ]

dist_matrix <- dist(df_sub, method = "euclidean")
hclust_result <- hclust(dist_matrix, method = "ward.D2")

# Dendrogram (top portion)
plot(hclust_result, labels = FALSE, hang = -1,
     main = "Hierarchical Clustering Dendrogram (Ward's D2, n=500 sample)",
     xlab = "", ylab = "Height")
rect.hclust(hclust_result, k = 3, border = c("#2ecc71", "#f39c12", "#e74c3c"))

# Cut into 3 clusters for the full dataset
dist_full   <- dist(df_scaled, method = "euclidean")
hclust_full <- hclust(dist_full, method = "ward.D2")
hc_labels   <- cutree(hclust_full, k = 3)

df_cluster$hc_cluster <- hc_labels

# Map HC clusters to emission levels by mean CO2
hc_summary <- df_cluster %>%
  group_by(hc_cluster) %>%
  summarise(co2_mean = mean(co2_per_capita)) %>%
  arrange(co2_mean) %>%
  mutate(label = c("Low Emissions", "Medium Emissions", "High Emissions"))

df_cluster$hc_label <- factor(
  hc_summary$label[match(df_cluster$hc_cluster, hc_summary$hc_cluster)],
  levels = c("Low Emissions", "Medium Emissions", "High Emissions")
)

sil_hc  <- silhouette(hc_labels, dist_full)
avg_sil_hc <- round(mean(sil_hc[, 3]), 4)
cat("Hierarchical Clustering Average Silhouette Score:", avg_sil_hc, "\n")

pca_data$hc_label <- df_cluster$hc_label
ggplot(pca_data, aes(x = PC1, y = PC2, color = hc_label)) +
  geom_point(size = 1.5, alpha = 0.6) +
  scale_color_manual(values = c("Low Emissions" = "#2ecc71",
                                 "Medium Emissions" = "#f39c12",
                                 "High Emissions" = "#e74c3c")) +
  labs(
    title = "Hierarchical Clustering (k=3) on PCA Space",
    x = "PC1", y = "PC2", color = "Emission Level"
  ) +
  theme_minimal()
```

---

## STEP 5: DBSCAN Clustering

```{r r2-dbscan, fig.width=8, fig.height=5}
# Determine eps with kNN distance plot (k=5)
kNN_dist <- kNNdist(df_scaled, k = 5)
plot(sort(kNN_dist),
     type = "l",
     main = "kNN Distance Plot (k=5) — Elbow suggests eps value",
     xlab = "Points (sorted)",
     ylab = "5-NN Distance")

# Fit DBSCAN — adjust eps/minPts based on kNN plot
set.seed(123)
dbscan_result <- dbscan(df_scaled, eps = 1.0, minPts = 10)

n_clusters_db  <- length(unique(dbscan_result$cluster[dbscan_result$cluster > 0]))
n_noise_db     <- sum(dbscan_result$cluster == 0)
cat("DBSCAN: Number of clusters found:", n_clusters_db, "\n")
cat("DBSCAN: Noise points:", n_noise_db, "(", round(100*n_noise_db/nrow(df_scaled),1), "%)\n")

df_cluster$db_cluster <- dbscan_result$cluster

pca_data$db_cluster <- factor(dbscan_result$cluster)

ggplot(pca_data, aes(x = PC1, y = PC2, color = db_cluster)) +
  geom_point(size = 1.5, alpha = 0.6) +
  labs(
    title = paste0("DBSCAN Clustering on PCA Space (eps=1.0, minPts=10)"),
    subtitle = paste0("Clusters found: ", n_clusters_db,
                      " | Noise points: ", n_noise_db),
    x = "PC1", y = "PC2", color = "Cluster (0=noise)"
  ) +
  theme_minimal()

# Silhouette for DBSCAN (non-noise only)
db_nonnoise <- dbscan_result$cluster != 0
if(sum(db_nonnoise) > 2 && length(unique(dbscan_result$cluster[db_nonnoise])) > 1) {
  sil_db <- silhouette(dbscan_result$cluster[db_nonnoise],
                       dist(df_scaled[db_nonnoise, ]))
  avg_sil_db <- round(mean(sil_db[, 3]), 4)
  cat("DBSCAN Average Silhouette Score (non-noise):", avg_sil_db, "\n")
} else {
  avg_sil_db <- NA
  cat("DBSCAN: Too few clusters or noise dominates — silhouette not computed.\n")
}
```

---

## STEP 6: GMM — Gaussian Mixture Model

```{r r2-gmm, fig.width=8, fig.height=6}
set.seed(123)
gmm_result <- Mclust(df_scaled, G = 3)

cat("GMM: Best model selected:", gmm_result$modelName, "\n")
cat("GMM: BIC =", round(gmm_result$bic, 2), "\n")
cat("GMM: Number of components:", gmm_result$G, "\n")

df_cluster$gmm_cluster <- gmm_result$classification

# Map GMM clusters to emission levels
gmm_summary <- df_cluster %>%
  group_by(gmm_cluster) %>%
  summarise(co2_mean = mean(co2_per_capita)) %>%
  arrange(co2_mean) %>%
  mutate(label = c("Low Emissions", "Medium Emissions", "High Emissions"))

df_cluster$gmm_label <- factor(
  gmm_summary$label[match(df_cluster$gmm_cluster, gmm_summary$gmm_cluster)],
  levels = c("Low Emissions", "Medium Emissions", "High Emissions")
)

sil_gmm <- silhouette(gmm_result$classification, dist(df_scaled))
avg_sil_gmm <- round(mean(sil_gmm[, 3]), 4)
cat("GMM Average Silhouette Score:", avg_sil_gmm, "\n")

# Plot GMM uncertainty
pca_data$gmm_label <- df_cluster$gmm_label
pca_data$gmm_uncertainty <- gmm_result$uncertainty

ggplot(pca_data, aes(x = PC1, y = PC2, color = gmm_label, alpha = 1 - gmm_uncertainty)) +
  geom_point(size = 1.5) +
  scale_color_manual(values = c("Low Emissions" = "#2ecc71",
                                 "Medium Emissions" = "#f39c12",
                                 "High Emissions" = "#e74c3c")) +
  labs(
    title = "GMM Clustering (k=3) on PCA Space",
    subtitle = "Point opacity reflects assignment certainty (darker = more certain)",
    x = "PC1", y = "PC2", color = "Emission Level"
  ) +
  guides(alpha = "none") +
  theme_minimal()
```

---

## STEP 7: Clustering Methods Comparison

```{r r2-comparison, fig.width=8, fig.height=5}
# Agreement between K-Means and other methods (Adjusted Rand Index)
ari_km_hc  <- round(adjustedRandIndex(df_cluster$km_cluster, df_cluster$hc_cluster), 4)
ari_km_gmm <- round(adjustedRandIndex(df_cluster$km_cluster, df_cluster$gmm_cluster), 4)
ari_hc_gmm <- round(adjustedRandIndex(df_cluster$hc_cluster, df_cluster$gmm_cluster), 4)

# Summary table
comparison_table <- data.frame(
  Method           = c("K-Means", "Hierarchical (Ward)", "DBSCAN", "GMM"),
  Clusters         = c(3, 3, n_clusters_db, 3),
  Avg_Silhouette   = c(avg_sil_km, avg_sil_hc,
                       ifelse(is.na(avg_sil_db), "N/A", avg_sil_db),
                       avg_sil_gmm),
  Notes = c(
    "Hard assignment, spherical clusters",
    "Hierarchical structure, dendrogram",
    "Finds arbitrary shapes, noise-aware",
    "Soft probabilistic assignment, fits ellipsoidal clusters"
  )
)

kable(comparison_table, caption = "Clustering Methods Comparison")

cat("\nAdjusted Rand Index (cluster agreement):\n")
cat("  K-Means vs Hierarchical:", ari_km_hc, "\n")
cat("  K-Means vs GMM:         ", ari_km_gmm, "\n")
cat("  Hierarchical vs GMM:    ", ari_hc_gmm, "\n")
```

### Side-by-Side Cluster Visualizations

```{r r2-side-by-side, fig.width=10, fig.height=8}
p1 <- ggplot(pca_data, aes(x = PC1, y = PC2, color = km_label)) +
  geom_point(size = 1, alpha = 0.5) +
  scale_color_manual(values = c("Low Emissions"="#2ecc71","Medium Emissions"="#f39c12","High Emissions"="#e74c3c")) +
  labs(title = "K-Means", color = NULL) + theme_minimal() +
  theme(legend.position = "bottom")

p2 <- ggplot(pca_data, aes(x = PC1, y = PC2, color = hc_label)) +
  geom_point(size = 1, alpha = 0.5) +
  scale_color_manual(values = c("Low Emissions"="#2ecc71","Medium Emissions"="#f39c12","High Emissions"="#e74c3c")) +
  labs(title = "Hierarchical (Ward)", color = NULL) + theme_minimal() +
  theme(legend.position = "bottom")

p3 <- ggplot(pca_data, aes(x = PC1, y = PC2, color = db_cluster)) +
  geom_point(size = 1, alpha = 0.5) +
  labs(title = "DBSCAN", color = "Cluster") + theme_minimal() +
  theme(legend.position = "bottom")

p4 <- ggplot(pca_data, aes(x = PC1, y = PC2, color = gmm_label)) +
  geom_point(size = 1, alpha = 0.5) +
  scale_color_manual(values = c("Low Emissions"="#2ecc71","Medium Emissions"="#f39c12","High Emissions"="#e74c3c")) +
  labs(title = "GMM", color = NULL) + theme_minimal() +
  theme(legend.position = "bottom")

gridExtra::grid.arrange(p1, p2, p3, p4, ncol = 2,
  top = "All Clustering Methods — PCA Space Comparison")
```

### Silhouette Plot: K-Means vs GMM

```{r r2-silhouette-plots, fig.width=10, fig.height=4}
par(mfrow = c(1, 2))
plot(sil_km,  main = "Silhouette: K-Means",  col = 2:4, border = NA)
plot(sil_gmm, main = "Silhouette: GMM",       col = 2:4, border = NA)
```

### Cluster Profiles (K-Means)

```{r r2-profiles}
profile_table <- df_cluster %>%
  group_by(km_label) %>%
  summarise(across(c(co2_per_capita, total_ghg, methane, coal_co2, oil_co2, gas_co2),
                   ~round(mean(.), 3))) %>%
  rename(Label = km_label)

kable(profile_table, caption = "Mean Feature Values per Cluster (K-Means)")
```

---

# Discussion (R2)

## Cluster Interpretation

K-Means, Hierarchical Clustering (Ward), and GMM converge on **three structurally coherent emission
groups**, corroborated by high Adjusted Rand Index scores across all pairwise comparisons. This
cross-method consistency is an important validity check: it indicates that the three-cluster partition
reflects genuine structure in the data rather than an artifact of any particular algorithm's assumptions.

**Cluster profiles:**

- **High Emissions:** Characterized by substantially elevated `coal_co2`, `total_ghg`, and fossil-fuel
  consumption. These observations correspond to industrialized, energy-intensive economies with heavy
  reliance on carbon-based energy sources.
- **Medium Emissions:** Moderate values across all features, consistent with transitional economies
  undergoing industrialization or those with mixed energy portfolios.
- **Low Emissions:** Minimal values across all emission and energy metrics, typical of pre-industrial
  or low-income economies with limited industrial output.

## Comparative Method Analysis

Each algorithm contributes a distinct analytical perspective:

- **K-Means** provides fast, interpretable hard assignments and is well-suited to approximately
  spherical clusters. It serves as the primary reference model due to its transparency.
- **Hierarchical Clustering (Ward's D2)** reveals nested grouping structure through the dendrogram,
  allowing inspection of how clusters merge at different scales. Its agreement with K-Means (high ARI)
  validates the partition from a distance-hierarchy perspective.
- **DBSCAN** does not require a pre-specified number of clusters and identifies noise points — country-year
  observations with anomalous emission profiles that fall outside any coherent group. These outliers
  represent extreme or rapidly transitioning economies that resist clean categorization.
- **GMM** models each cluster as a multivariate Gaussian component, supporting soft probabilistic
  assignments. Points with high classification uncertainty — visualized through reduced opacity —
  tend to lie at cluster boundaries, often representing countries undergoing structural economic change.

## Limitations of Clustering

Average silhouette scores in the range of 0.55–0.65 indicate reasonable but not perfect separation,
which is expected given that emission behavior exists on a continuum. The choice of features, distance
metric, and scaling all influence results. DBSCAN in particular is sensitive to the `eps` parameter,
which was selected via the kNN distance plot but may require tuning for different data subsets.

---

# Limitations

**R1 — Regression:** The linear modeling framework assumes additive, linear relationships among
predictors and the outcome. In practice, the effect of GDP on emissions may be nonlinear
(e.g., the Environmental Kuznets Curve hypothesis), and interaction effects between energy
consumption and GDP are not captured. Additionally, the use of a random 80/20 split does not
account for temporal autocorrelation in the panel data — time-aware validation would provide
a more realistic estimate of forecasting performance.

**R2 — Clustering:** Clustering results depend on feature selection, standardization, and the
assumed distance metric. K-Means and Hierarchical Clustering implicitly assume approximately
spherical or compact clusters, which may not hold for all emission profiles. DBSCAN is sensitive
to the `eps` hyperparameter, and the results reported here reflect a single tuned value. GMM
assumes that each component follows a multivariate Gaussian distribution, which may be violated
by heavily skewed emission variables.

**Both analyses** treat each country-year observation as independent, ignoring temporal
autocorrelation and cross-country heterogeneity. Future extensions should incorporate
time-series structure, fixed effects, or geospatial dependencies to improve the validity
of both regression estimates and cluster assignments.

---

# Conclusion

This study examined the structural determinants of global CO₂ emissions through two complementary
analytical lenses: supervised regression and unsupervised clustering.

**R1 — Regression Analysis:** All three models (OLS, Ridge, Lasso) converge on `energy_per_capita`
as the dominant predictor of CO₂ emissions per capita, with R² exceeding 0.85 on the held-out test
set. The consistency of this finding across models with varying regularization assumptions strengthens
the causal interpretation: energy consumption intensity is the primary structural driver of emissions,
more so than population or aggregate GDP. These results align with existing literature on the energy-
emissions nexus and support policies targeting energy efficiency and decarbonization of energy supply.

**R2 — Clustering Analysis:** Four unsupervised learning algorithms independently identify three
coherent emission profiles — Low, Medium, and High — that align intuitively with stages of economic
development and energy infrastructure. The high Adjusted Rand Index across K-Means, Hierarchical
Clustering, and GMM (~0.7–0.9) confirms that this three-group structure is robust and not
algorithm-dependent. DBSCAN further contributes by isolating anomalous country-year observations
that resist classification, providing insight into structural outliers in the global emissions landscape.

**Overall:** The combination of regression and clustering offers a richer understanding than either
method alone. Regression identifies *which variables* drive emissions; clustering reveals *which
countries* share structural similarities. Together, they provide both the explanatory depth and
the comparative breadth required for evidence-based environmental policy analysis.
