---
title: "CO2 Emissions Analysis: Regression and Clustering"
output: 
  pdf_document:
    latex_engine: xelatex
author: "Group-9"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#  Introduction

This analysis investigates global CO2 emissions using statistical and machine learning techniques.
Two main research questions are addressed:

1. **R1:** What factors influence CO2 emissions per capita?
2. **R2:** Can countries be grouped based on emission patterns?

---

#  Data Preparation

```{r}
# Purpose: Load all packages needed for data prep, modeling, diagnostics, and plotting.
library(tidyverse)
library(glmnet)
library(caret)
library(factoextra)
library(cluster)
library(corrplot)
library(car)
library(knitr)

# Load dataset
df <- read.csv("cleaned_co2_data_20vars.csv")

# Display structure
str(df)
```


# Exploratory Data Analysis (EDA)

```{r}

# Purpose: Quick EDA to visualize GDP vs CO2-per-capita relationship.

ggplot(df, aes(x = gdp, y = co2_per_capita)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_smooth(method = "lm", color = "red") +
  labs(
    title = "Relationship Between GDP and CO2 Emissions",
    subtitle = "Higher GDP is generally associated with higher emissions",
    x = "GDP",
    y = "CO2 Emissions per Capita"
  ) +
  theme_minimal()
```

---

#  Research Question 1: Regression Analysis

## Purpose

To understand how GDP, population, and energy consumption affect CO₂ emissions.

```{r}

# Purpose: Build R1 modeling dataset (target + selected predictors), then drop NAs.

df_model <- df %>%
  select(co2_per_capita, gdp, population, energy_per_capita) %>%
  na.omit()

# Purpose: Create an 80/20 train-test split for out-of-sample evaluation.
set.seed(123)
trainIndex <- createDataPartition(df_model$co2_per_capita, p = 0.8, list = FALSE)

train_data <- df_model[trainIndex, ]
test_data  <- df_model[-trainIndex, ]
```


```{r}

# Purpose: Inspect pairwise correlation and potential multicollinearity in R1 variables.

# Loaded in setup chunk
# library(corrplot)

cor_matrix <- cor(df_model)
corrplot(cor_matrix, method = "color", tl.cex = 0.8)
```
---

##  Model 1: OLS Regression

```{r}
# Purpose: Fit OLS baseline model and generate test predictions.
ols_model <- lm(co2_per_capita ~ gdp + population + energy_per_capita, data = train_data)

# Predictions
ols_pred <- predict(ols_model, newdata = test_data)

# Evaluation metrics
ols_rmse <- sqrt(mean((test_data$co2_per_capita - ols_pred)^2))
ols_r2   <- cor(test_data$co2_per_capita, ols_pred)^2
ols_mae  <- mean(abs(test_data$co2_per_capita - ols_pred))


# Purpose: Show coefficient estimates/significance for interpretation.

summary(ols_model)


# Purpose: Check multicollinearity using VIF (required R1 diagnostic).

# Loaded in setup chunk
# library(car)
vif(ols_model)
```

---

##  Model 2: Ridge Regression

```{r}
# Purpose: Convert data to matrix format for regularized models (glmnet).
X_train <- model.matrix(co2_per_capita ~ gdp + population + energy_per_capita, train_data)[,-1]
y_train <- train_data$co2_per_capita

X_test <- model.matrix(co2_per_capita ~ gdp + population + energy_per_capita, test_data)[,-1]
y_test <- test_data$co2_per_capita

# Purpose: Train Ridge regression with CV-selected lambda.
ridge_model <- cv.glmnet(X_train, y_train, alpha = 0)

# Predictions
ridge_pred <- predict(ridge_model, s = "lambda.min", newx = X_test)

# Metrics
ridge_rmse <- sqrt(mean((y_test - ridge_pred)^2))
ridge_r2   <- cor(y_test, ridge_pred)^2
ridge_mae  <- mean(abs(y_test - ridge_pred))
```

---

##  Model 3: Lasso Regression

```{r}
# Purpose: Train Lasso regression with CV-selected lambda.
lasso_model <- cv.glmnet(X_train, y_train, alpha = 1)

# Predictions
lasso_pred <- predict(lasso_model, s = "lambda.min", newx = X_test)

# Metrics
lasso_rmse <- sqrt(mean((y_test - lasso_pred)^2))
lasso_r2   <- cor(y_test, lasso_pred)^2
lasso_mae  <- mean(abs(y_test - lasso_pred))

# Feature Importance from Lasso

coef(lasso_model, s = "lambda.min")
```

---

##  Model Comparison

```{r}
results <- data.frame(
  Model = c("OLS", "Ridge", "Lasso"),
  RMSE  = c(ols_rmse, ridge_rmse, lasso_rmse),
  R2    = c(ols_r2, ridge_r2, lasso_r2),
  MAE   = c(ols_mae, ridge_mae, lasso_mae)
)

kable(results, caption = "Model Comparison (RMSE, R², MAE)")

cat("Best Model Based on RMSE:", results$Model[which.min(results$RMSE)])
```

---

##  Visualization: Model Performance

```{r}
# Purpose: Plot actual vs predicted for visual model quality check.
plot_df <- data.frame(actual = y_test, predicted = as.numeric(lasso_pred))

# Plot
ggplot(plot_df, aes(x = actual, y = predicted)) +
  geom_point(color = "blue", alpha = 0.6) +
  geom_abline(slope = 1, intercept = 0, color = "red", linetype = "dashed") +
  labs(
    title = "Actual vs Predicted CO2 Emissions (Lasso Model)",
    x = "Actual CO2",
    y = "Predicted CO2",
    caption = "Red dashed line indicates perfect predictions"
  ) +
  theme_minimal()
```
```{r}

# Purpose: Residual diagnostics for OLS assumptions.

par(mfrow=c(2,2))
plot(ols_model)
```

---

##  Interpretation (R1)

Lasso regression slightly outperformed both OLS and Ridge models in terms of R² and RMSE.

This indicates that Lasso’s regularization helps reduce overfitting and improves generalization performance.

The regression coefficients from the OLS model show that energy consumption per capita has a strong positive 
and statistically significant effect on CO₂ emissions. GDP also contributes positively, while population 
is not statistically significant in this model.

This suggests that energy usage is the primary driver of emissions rather than population size alone.

Additionally, Lasso regression highlights the most important predictors by shrinking less relevant variablestoward zero, reinforcing the importance of energy-related variables in explaining emissions.

The consistency across OLS, Ridge, and Lasso models suggests a robust linear relationship between energy consumption and CO₂ emissions.
---

# Research Question 2: Clustering Analysis

## Purpose

To group countries based on emission characteristics and identify patterns in CO₂ emissions.

**Important clarification (R2):** Clustering is done on **country-year observations** (not only country-level averages) so temporal variation is preserved.

```{r}

# STEP 1: Feature Selection and Data Scaling
# Purpose: Select relevant emission-related variables

df_cluster <- df %>%
  select(co2_per_capita, total_ghg, methane, coal_co2, oil_co2, gas_co2) %>%
  na.omit()

# Standardize features (important for clustering)
df_scaled <- scale(df_cluster)
```

---

```{r}

# STEP 2: PCA (Dimensionality Reduction)
# Purpose: Understand variance and visualize clusters

pca <- prcomp(df_scaled)

# Scree plot
fviz_eig(pca) +
  labs(
    title = "Scree Plot: Variance Explained by Principal Components",
    x = "Principal Components",
    y = "Percentage of Variance Explained"
  ) +
  theme_minimal()
```

---

```{r}

# STEP 3: Determine Optimal Number of Clusters

# Elbow Method
fviz_nbclust(df_scaled, kmeans, method = "wss") +
  labs(
    title = "Elbow Method for Optimal Number of Clusters",
    x = "Number of Clusters (k)",
    y = "Within-Cluster Sum of Squares"
  ) +
  theme_minimal()

# Silhouette Method
fviz_nbclust(df_scaled, kmeans, method = "silhouette") +
  labs(
    title = "Optimal Clusters using Silhouette Method"
  ) +
  theme_minimal()
```

---

```{r}

# STEP 4: Compare Different K Values
# Purpose: Validate best number of clusters

for(k in 2:4){
  km <- kmeans(df_scaled, centers = k, nstart = 25)
  sil <- silhouette(km$cluster, dist(df_scaled))
  cat("k =", k, "Average Silhouette =", mean(sil[,3]), "\n")
}
```

---

```{r}

# STEP 5: Final K-Means Model (k = 3)

set.seed(123)
kmeans_result <- kmeans(df_scaled, centers = 3, nstart = 25)

df_cluster$cluster <- kmeans_result$cluster
```

---

```{r}

# STEP 6: Label Clusters (Low / Medium / High Emissions)
# Purpose: Make clusters interpretable

cluster_summary <- df_cluster %>%
  group_by(cluster) %>%
  summarise(co2_mean = mean(co2_per_capita))

# Sort clusters by emission level
cluster_summary <- cluster_summary %>%
  arrange(co2_mean)

cluster_summary

# Assign labels
labels <- c("Low Emissions", "Medium Emissions", "High Emissions")

cluster_summary$label <- labels

# Map labels back to dataset
df_cluster$cluster_label <- factor(
  cluster_summary$label[match(df_cluster$cluster, cluster_summary$cluster)],
  levels = labels
)
```

---

```{r}

# STEP 7: PCA Visualization with Meaningful Labels

pca_data <- as.data.frame(pca$x)
pca_data$cluster_label <- df_cluster$cluster_label

ggplot(pca_data, aes(x = PC1, y = PC2, color = cluster_label)) +
  geom_point(size = 2, alpha = 0.7) +
  labs(
    title = "Country Clusters Based on CO2 Emissions",
    subtitle = "Clusters categorized as Low, Medium, and High Emissions",
    x = "Principal Component 1",
    y = "Principal Component 2",
    color = "Emission Level"
  ) +
  theme_minimal()
```

---

```{r}

# STEP 8: Silhouette Analysis
# Purpose: Evaluate clustering quality

library(cluster)
library(corrplot)
library(car)

sil <- silhouette(kmeans_result$cluster, dist(df_scaled))

# Average silhouette score
avg_sil <- mean(sil[,3])
cat("Average Silhouette Score:", avg_sil, "\n")

# Silhouette plot
fviz_silhouette(sil) +
  labs(
    title = "Silhouette Plot for K-Means Clustering",
    subtitle = "Higher values indicate better-defined clusters",
    x = "Cluster",
    y = "Silhouette Width"
  ) +
  theme_minimal()
```

---

```{r}

# STEP 9: Cluster Summary (Interpretation Table)

df_cluster %>%
  group_by(cluster_label) %>%
  summarise(across(everything(), mean))
```

---

# 🧠 Interpretation (R2)

The clustering results reveal clear segmentation of countries into three emission categories.

The "High Emissions" cluster is characterized by significantly higher energy consumption and greenhouse gas levels, 
indicating industrialized or highly developed economies.

The "Medium Emissions" cluster represents transitional economies with moderate energy usage and emissions.

The "Low Emissions" cluster consists of countries with minimal industrial activity and lower environmental impact.

The average silhouette score (~0.58) indicates reasonably well-separated clusters. However, the presence of a cluster 
with lower silhouette values suggests some overlap between emission groups, particularly among medium-emission countries.

This indicates that while clustering captures general patterns effectively, emission behavior exists on a spectrum 
rather than in strictly separated categories.

---

# Limitations

This analysis assumes linear relationships in regression models, which may not fully capture 
complex interactions between variables.

Additionally, clustering results depend on feature selection and scaling, and alternative methods 
such as hierarchical clustering, DBSCAN, or GMM may produce different groupings.

Future work could include non-linear modeling approaches and time-series analysis to better understand 
temporal emission patterns.

---

# Conclusion

This analysis demonstrates that energy consumption is the dominant driver of CO₂ emissions.

Regression models show strong predictive performance, with Lasso providing slightly better generalization due to regularization.

Clustering reveals meaningful groupings of countries into low, medium, and high emission categories, highlighting global disparities in environmental impact.

Overall, this study demonstrates the effectiveness of combining statistical modeling and unsupervised learning to analyze complex global environmental data.
